---
title: "Lab 4: Radical_Right"
author: "Ross Campbell"
date: '2025-09-01'
output: word_document
---

```{r setup, include=FALSE}
ess <- foreign::read.dta("https://raw.github.com/drcampbell1/Participation/master/data/ess.dta", convert.factors = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(foreign)
library(tidyverse)
library(kableExtra)
```

In the lecture, we introduced theories about who votes for the radical right and took a look at some of the trends. In this lab, we're going to test some aspects of those theories with evidence from the European Social Survey.

```{r, echo = FALSE, results='asis'}

ess %>% 
  filter(!is.na(right)) %>% 
  count(right) %>%
  mutate("%" = round(n/sum(n) * 100, digits=1)) %>%
  knitr::kable("pandoc", caption = "Voted for the Radical Right in European Democracies",
               col.names = c('Voted Radical Right', 'N', '%'), align="lccc")
```

How does this vary by country?

```{r, echo = FALSE, results='asis'}

ess %>% 
  group_by(country) %>% 
  filter(!is.na(right)) %>%
     count(right) %>% 
  mutate("%" = round(n/sum(n) * 100, digits=1)) %>% filter(!right =="did not vote") %>% 
     knitr::kable("pandoc", caption = "Comparing Voting for Radical Right in European Democracies",
     col.names = c('Country', 'Voted Radical Right', 'N', '%'), align="cccc")
```

Let's graph this to make the findings stand out a bit better

```{r, echo = FALSE}
ess %>% 
  group_by(country) %>% 
  filter(!is.na(right)) %>%
  count(right) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!right=="did not vote") %>%
  ggplot(aes(x=reorder(country, -prop), y=prop)) +
  geom_col()+
  labs(x="", y="", title="Figure 1: Voted for Radical Right Party by Country", caption="ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 30, hjust =1))

```

A lot of studies connect voting for the radical right to age. One of the assumptions is that radical right parties successfully mobilise younger, more disaffected sections of the population. Let's test this by comparing how the average age of radical right voters compares with the average age of the population overall.

```{r, include = FALSE}

a <- ess %>% 
  group_by(country) %>% 
             summarise(mean= mean(age, na.rm=T))
b <- ess %>% 
  group_by(country, right) %>% 
             summarise(avg_age_vt_rr= mean(age, na.rm=T)) %>% 
             filter(!is.na(right), !right=="did not vote")
```

```{r, echo = FALSE, results = 'asis'}
left_join(a, b, by = "country") %>% 
  select(-right) %>%
          knitr::kable("pandoc", caption = "Average Age of Population and Voters of Radical Right Parties in European Democracies",
          col.names = c('Country', 
                        'Average Age Population', 
                        'Average Age Voter Radical Right Party'), 
              digits =1, align="cccc")
```

What does this tell us about the age profile of voters for radical right parties?

A similarly important claim is about the gender profile of radical right voters. Again, the assumption is that radical right voters are disproportionately male. In other words, males are *over represented* amongst the radical right electorate.

```{r, include = FALSE}

c <- ess %>%
  group_by(country) %>%
  filter(!is.na(gender))%>%
  count(gender) %>%
  mutate(perc_pop=n/sum(n)*100)


d <- ess %>%
  filter(!is.na(right)) %>%
  group_by(country, gender) %>% 
  filter(!right=="did not vote") %>% count(right) %>% 
  group_by(country) %>% 
  mutate(perc_vt_pp=n/sum(n)*100) %>% 
  select(-right)
```

```{r, echo = FALSE, results='asis'}

left_join(c, d, by=c("country", "gender")) %>% select(-n.x, -n.y) %>%
  kbl("pandoc", caption = "Gender Balance in Population and Voters of Radical Right Parties in European Democracies",col.names = c('Country', 'Gender', 'Percentage Population', 'Percentage Voter of Radical Right Political Party'),
               digits =1, align="cccc") 
```

```{r, include = FALSE}
rm(a,b,c,d)
```

Let's look at Education

```{r, echo = FALSE}
ess %>% 
  group_by(educat) %>% 
  filter(!is.na(right), !is.na(educat)) %>%
  count(right) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!right=="did not vote") %>%
  ggplot(aes(x=educat, y=prop)) +
  geom_col()+
  labs(x="", 
       y="", 
       title="Figure 2: Voting for the Radical Right by Education", 
       caption="Source: ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 30, hjust =1))
``` 

```{r, echo = FALSE}

ess %>% 
  group_by(educat, country) %>% 
filter(!is.na(right), !is.na(educat)) %>%
  count(right) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!right=="did not vote") %>%
  ggplot(aes(x=educat, y=prop)) +
  geom_col()+
  facet_wrap(~country, nrow=3)+
  labs(x="", 
       y="", 
       title="Figure 3: Voting for the Radical\nRight by Education and Country", 
       caption="Source: ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 30, hjust =1))
```

Let's look at what voters of the radical right think about government?

Do voters of the radical right trust government?

```{r, echo = FALSE}
ess %>% 
  group_by(ptrust) %>% 
  filter(!is.na(right), !is.na(ptrust)) %>%
  count(right) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!right=="did not vote") %>%
  ggplot(aes(x=ptrust, y=prop)) +
  geom_col()+
  labs(x="", y="", title="Figure 4: Political Trust\nand Voting for the Radical Right", caption="Source: ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 30, hjust =1))
```

```{r, echo = FALSE}
ess %>% 
  group_by(ptrust, country) %>% 
  filter(!is.na(right), !is.na(ptrust)) %>%
  count(right) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!right=="did not vote") %>%
  ggplot(aes(x=ptrust, y=prop)) +
  geom_col()+
  facet_wrap(~country, nrow=3)+
  labs(x="",
       y="", 
       title="Figure 5: Political Trust\nand Voting for the Radical Right", 
       caption="Source: ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 30, hjust =1))
``` 

Attitudes to Immigrants of a Different Ethnicity

```{r, echo = FALSE}
ess %>% 
  group_by(imdfetn) %>% 
  filter(!is.na(right), !is.na(imdfetn)) %>%
  count(right) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!right=="did not vote") %>%
  ggplot(aes(x=imdfetn, y=prop)) +
  geom_col()+
  labs(x="", 
       y="", 
       title="Figure 6: Attitudes to Immigration\nand Voting for the Radical Right",
       subtitle = "(different ethnicity)",
       caption="ESS: 2002-2016")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()
```

```{r, echo = FALSE}

ess %>% 
  group_by(imdfetn, country) %>% 
  filter(!is.na(right), !is.na(imdfetn)) %>%
  count(right) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!right=="did not vote") %>%
  ggplot(aes(x=imdfetn, y=prop)) +
  geom_col()+
  facet_wrap(~country, nrow=3)+
  labs(x="", 
       y="", 
       title="Figure 7: Attitudes to Immigration\nand Voting for the Radical Right", 
       caption="ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()
```
 
 What can we conclude about voters for the radical right?