---
title: "Lab 1: Introduction"
output: word_document
author: "Ross Campbell PhD"
date: "2025-08-28"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(foreign)
library(tidyverse)
library(tidytext)
library("kableExtra")
ess <- foreign::read.dta("https://raw.github.com/drcampbell1/Participation/master/data/ess.dta", convert.factors = TRUE)

```

## Introduction to the data

In the labs we'll be using the ESS. This is the European Social Survey - a big cross-sectional survey conducted in 30 countries in Europe.


## Coverage
To make the file more manageable, I've cut it down quite a bit.

### Basic Terminology

1. Dependent variable: something we are looking to explain, e.g. one (or more) forms of participation:
2. Independent variable : something that influences our dependent variable (age, for example, may influence voting).
3. Observation: someone we ask about their participation and age.
 
### Overall, then, this file contains :
8 different forms of participation; 8 different variables that may influence participation (or 9 if we include country); and And we have 78473 observations - or 78473 people gave us answers to our questions.

```{r, echo = FALSE, results = 'asis'}
ess %>% 
  count(country) %>% 
  knitr::kable("pandoc", caption = "Coverage of European Social Survey",
               col.names = c("Country", "Number of Observations"), align = "cc")
```

For each country, we also have data over time:

```{r, echo = FALSE, results = 'asis'}
ess %>% 
  count(year) %>% 
  knitr::kable("pandoc", caption = "Coverage in time of European Social Survey",
               col.names = c("Year", "Observations"), align = "cc")
```

Let's finish today by looking at some forms of participation:

How many people said they voted at the last national election?
```{r, echo = FALSE, results = 'asis'}
ess %>% 
  filter(!is.na(vote)) %>% 
  count(vote) %>%
  mutate('%' = round(n/sum(n)*100, digits=1)) %>% 
    knitr::kable("pandoc", caption = "Electoral Participation",
               col.names = c("Voted", "Observations", "%"), align = "ccc")

```

#Let's look at contacting politicians
```{r, echo = FALSE}
ess %>% 
  filter(!is.na(contact)) %>% count(contact) %>% 
  mutate('%' = round(n/sum(n)*100, digits=1)) %>% 
  knitr::kable("pandoc", caption = "Contacting",
               col.names = c("Contacted Politicians", "Observations", "%"), align = "ccc")
```

#Lets look at demonstrating

```{r, echo = FALSE}
ess %>% 
  filter(!is.na(demo)) %>% count(demo) %>% 
  mutate('%' = round(n/sum(n)*100, digits=1)) %>% 
  knitr::kable("pandoc", caption = "Demonstrating",
               col.names = c("Have Demonstrated", "Observations", "%"), align = "ccc")
```

# Let's look at voting for the radical right
```{r, echo = FALSE}
ess %>% 
  filter(!is.na(right)) %>% count(right) %>% 
  mutate('%' = round(n/sum(n)*100, digits=1)) %>% 
  knitr::kable("pandoc", caption = "Voted Radical Right",
               col.names = c("Voted Radical Right", "Observations", "%"), align = "ccc")
```


What if we compared a lot of forms across countries?
```{r, echo = FALSE, fig.height=25, fig.width=35}
ess %>% 
  pivot_longer((vote:petit), 
               names_to = "mode", 
               values_to = "value") %>% 
  group_by(country, mode) %>%
  filter(!is.na(value)) %>% 
  count(value) %>% 
  mutate(percent = n/sum(n)*100) %>% 
  filter(value == str_remove(value, "not")) %>% 
  mutate(value=str_replace(value, "have", ""), 
         value = reorder_within(value, percent, country)) %>% 
  ggplot(aes(reorder(value, percent), percent))+
  geom_col(fill= "steelblue")+
  facet_wrap(~country, scales = "free")+
  coord_flip()+
  theme_minimal()+
  scale_x_reordered()+
  theme(axis.text.x = element_text(size = 46),
              axis.text.y = element_text(size = 46),
        plot.title = element_text(size = 46),
        plot.subtitle = element_text(size=46),
        plot.caption = element_text(size = 40))+
  labs(x = "", y = "%", 
       title = "Figure 1:Comparing Political\nParticipation in European Democracies",
       subtitle = "(Political Styles that Differ)",
       caption = "Source: ESS 2002-2018")
```

