---
title: "Lab 2: Voting"
output: word_document
author: "Ross Campbell PhD"
date: "2025-08-28"
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(foreign)
library(tidyverse)
library(tidytext)
library("kableExtra")
ess <- foreign::read.dta("https://raw.github.com/drcampbell1/Participation/master/data/ess.dta", convert.factors = TRUE)

```

### Exercise 1: Let's look at voting#
 Let's ask a simple question: do people vote?

```{r, echo = FALSE, results='asis'}
ess %>% 
  filter(!is.na(vote)) %>% 
  count(vote) %>%
  mutate("%" = round(n/sum(n) * 100, digits=1)) %>%
  knitr::kable("pandoc",caption = "Electoral Turnout in European Democracies",
  col.names = c('voted', 'Observations', '%'), align="lccc") 
```

### How does this vary by country?#

```{r, echo = FALSE, results = 'asis'}
ess %>% 
  group_by(country) %>% 
  filter(!is.na(vote)) %>%
  count(vote) %>% 
  mutate("%" = round(n/sum(n) * 100, digits=1)) %>%
  knitr::kable("pandoc", caption = "Electoral Turnout in European Democracies",
  col.names = c('Country', 'Voted', 'N', '%'), align="cccc") 
```

### We can present this graphically to make it clearer#

```{r, echo = FALSE, fig.height=25, fig.width=35}
ess %>% 
  group_by(country) %>% 
  filter(!is.na(vote)) %>%
  count(vote) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!vote=="did not vote") %>%
  ggplot(aes(x=reorder(country, -prop), y=prop)) +
  geom_bar(stat="identity")+
  labs(x="", y="", title="Figure 1: Turnout by Country", caption="Source: ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
    theme(axis.text.x = element_text(size = 46),
              axis.text.y = element_text(size = 46),
        plot.title = element_text(size = 46),
        plot.subtitle = element_text(size=46),
        plot.caption = element_text(size = 40))
```

### What's changed over time? 

```{r, echo=FALSE, fig.height=25, fig.width=35}
ess %>% 
  group_by(year, country) %>% 
  filter(!is.na(vote)) %>%
  count(vote) %>% mutate(prop=prop.table(n*100)) %>%
  filter(!vote=="did not vote") %>%
  ggplot(aes(year, prop))+
  geom_line()+
  facet_wrap(~country, nrow = 3)+
  labs(x="", y="", title="Figure 2: Turnout by Country Over Time", caption="ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  scale_x_continuous(breaks = c(2002, 2006, 2010, 2014, 2018))+
  theme(axis.text.x = element_text(size = 46),
              axis.text.y = element_text(size = 46),
        plot.title = element_text(size = 46),
        plot.subtitle = element_text(size=46),
        plot.caption = element_text(size = 40),
        strip.text = element_text(size = 35))
```

### Exercise 2: Let's test our theories
Q1: Do Socio-Economic Resources Matter. What about Education?

```{r, echo = FALSE, results = 'asis'}
ess %>% 
  group_by(educat) %>% 
  filter(!is.na(vote), !is.na(educat)) %>%
  count(vote) %>% 
  mutate("%" =n/sum(n)*100) %>%
  knitr::kable("pandoc", caption = "Education and Electoral Turnout",
  col.names = c('Education', 'Voted', 'N', '%'), align="ccccc", digits = 1) 

```

We can graph the relationship

```{r, echo=FALSE, fig.height=25, fig.width=35}
ess %>% 
  group_by(educat) %>% 
  filter(!is.na(vote), !is.na(educat)) %>%
  count(vote) %>% 
  mutate(prop=prop.table(n*100)) %>% 
  mutate("%" =prop*100) %>%
  ggplot(aes(x=educat, y=prop, fill=vote))+
  labs(x="", y="", title="Figure 2: Turnout by Education", caption="ESS 2002-2018")+
  scale_fill_manual(values=c("darkred", "steelblue"))+
  geom_bar(stat="identity", position="dodge")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 40),
        legend.position = "bottom",
        axis.text.x = element_text(size = 46),
        axis.text.y = element_text(size = 46),
        plot.title = element_text(size = 46),
        plot.subtitle = element_text(size=46),
        plot.caption = element_text(size = 40))
```

### Turnout by Education and Country 
```{r, echo = FALSE, fig.height=25, fig.width=35}
ess %>% 
  group_by(educat, country) %>% 
  filter(!is.na(vote), !is.na(educat)) %>%
  count(vote) %>% 
  mutate(prop=prop.table(n*100)) %>% 
  mutate("%" =prop*100) %>%
  ggplot(aes(x=educat, y=prop, fill = vote))+
  labs(x="", 
       y="", title="Figure 3: Turnout by Education and Country", 
       caption="Source: ESS 2002-2018")+
  geom_bar(stat = "identity", position = "dodge")+
  scale_fill_manual(values=c("darkred", "steelblue"))+
  facet_wrap(~country, nrow=3)+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 40),
        legend.position = "bottom",
        axis.text.x = element_text(size = 46),
        axis.text.y = element_text(size = 46),
        plot.title = element_text(size = 46),
        plot.subtitle = element_text(size=46),
        plot.caption = element_text(size = 40),
        strip.text = element_text(size = 35))
```

### Q2: Does Rationality Influence Turnout?

```{r, echo = FALSE, results = 'asis'}

ess %>% 
  group_by(satecon) %>% 
  filter(!is.na(vote), !is.na(satecon)) %>%
  count(vote) %>% 
  mutate(prop= n /sum(n)*100) %>%
  knitr::kable("pandoc", caption = "Perceptions of the Economy and Electoral Turnout",
  col.names=c('Economic Perceptions', 'Voted', 'N', '%'), align="ccccc", digits=1)
```

### We can graph this relationship
```{r, echo = FALSE, fig.height=25, fig.width=35}
ess %>% 
  group_by(satecon) %>% 
  filter(!is.na(vote), !is.na(satecon)) %>%
  count(vote) %>% 
  mutate(prop=prop.table(n*100)) %>% 
  mutate("%" =prop*100) %>%
  ggplot(aes(x=satecon, y=prop, fill = vote))+
  labs(x="", y="", title="Figure 4: Turnout by Perceptions of the Economy", caption="Source: ESS 2002-2018")+
  geom_bar(stat = "identity", position = "dodge")+
  scale_fill_manual(values=c("darkred", "steelblue"))+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  scale_x_discrete(labels=c("low" = "low", "medium" = "medium", "high" = "high"))+
    theme(legend.title = element_blank(),
        legend.text = element_text(size = 40),
        legend.position = "bottom",
        axis.text.x = element_text(size = 46),
        axis.text.y = element_text(size = 46),
        plot.title = element_text(size = 46),
        plot.subtitle = element_text(size=46),
        plot.caption = element_text(size = 40))
```

### Does it differ between the countries?

```{r, echo = FALSE, fig.height=25, fig.width=35}

ess %>% group_by(satecon, country) %>% filter(!is.na(vote), !is.na(satecon)) %>%
  count(vote) %>% 
  mutate(prop=prop.table(n*100)) %>% 
  mutate("%" =prop*100) %>%
  ggplot(aes(x=satecon, y=prop, fill = vote))+
  labs(x="", y="", title="Figure 5: Turnout by Economic Satisfaction and Country", caption="ESS 2002-2018")+
  geom_bar(stat = "identity", position = "dodge")+
  scale_fill_manual(values=c("darkred", "steelblue"))+
  facet_wrap(~country, nrow=3)+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  theme(legend.title = element_blank())+
  theme(legend.position = "bottom")+
  scale_x_discrete(labels=c("low" = "low", "medium" = "medium", "high" = "high"))+
  theme(legend.title = element_blank(),
        legend.text = element_text(size = 40),
        legend.position = "bottom",
        axis.text.x = element_text(size = 46),
        axis.text.y = element_text(size = 46),
        plot.title = element_text(size = 46),
        plot.subtitle = element_text(size=46),
        plot.caption = element_text(size = 40))
```

### Q3 Does history matter? Does the legacy of post-communism influence participation?

```{r, echo = FALSE, results = 'asis'}
ess %>% 
  filter(!is.na(vote)) %>% 
  mutate(post_communist = if_else(country == "Poland", "post-communist", "not post-communist")) %>% 
  group_by(post_communist) %>% 
  count(vote) %>% 
  mutate(percent = round(n /sum(n)*100, digits = 1)) %>% 
  filter(vote == "voted")%>%
  knitr::kable("pandoc", caption = "The Legacy of Communism and Electoral Turnout",
      col.names=c('Country', 'Voted', 'N', '%'), align="ccccc", digits=1)
```

### Q4 Representative vs. Direct Democracy

```{r, echo = FALSE, results = 'asis'}
ess %>% 
  filter(!is.na(vote)) %>% 
  mutate(direct_democracy = if_else(country == "Switzerland", "Direct Democracy", "Representative Democracy")) %>% 
  group_by(direct_democracy) %>% 
  count(vote) %>% 
  mutate(percent = round(n /sum(n)*100, digits = 1)) %>% 
  filter(vote == "voted") %>%
  knitr::kable("pandoc", caption = "Representative vs. Direct Democracy and Electoral Turnout",
      col.names=c('Type of Democracy', 'Voted', 'N', '%'), align="ccccc", digits=1)
```

### Bonus Question: Let's ask if young people vote. 
We can do this by comparing the percentage of young people found in the sample with the percentage found amongst the voters?

```{r, echo = FALSE, results = 'asis'}

a <- ess %>% 
filter(!is.na(age)) %>% 
mutate(generation = if_else(age >= 18 & age <= 25, "18-25", "26+")) %>% 
count(generation) %>% 
mutate(percent = round(n /sum(n)*100, digits = 1)) 

b <- ess %>% 
filter(!is.na(age)) %>% 
mutate(generation = if_else(age >= 18 & age <= 25, "18-25", "26+")) %>% 
filter(vote == "voted") %>% 
group_by(generation) %>% 
count(vote) %>% 
ungroup() %>% 
mutate(percent = round(n/sum(n)*100, digits = 1))

left_join(a, b, by = "generation") %>% 
select(-c(n.x, n.y, vote)) %>% 
knitr::kable(caption = "Percentage of Age within Sample and Percentage of Age Cohort amongst Voters", col.names = c('Generation', 'Percentage Within Sample', 'Percentage Within Voters'), 
             align="ccc") 
rm(a,b)
```

### Three questions to conclude with:

So, thinking over the theories we've look at: which ones work?

Do all work equally well in all countries?

Does it matter that young people seem less likely to vote?










