---
title: "Lab 6: Demonstrating"
author: "Ross Campbell"
date: "2025-09-17"
output: word_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(foreign)
library(tidyverse)
library(tidytext)
library("kableExtra")
ess <- foreign::read.dta("https://raw.github.com/drcampbell1/Participation/master/data/ess.dta", convert.factors = TRUE)
```

In the lecture I argued that demonstrating has expanded and diversified as a form of participation. Whilst it was once relatively uncommon, it has now grown in frequency. And whilst it was once the preserve of younger and more ideologically extreme males, it is now used by a wide demographic.

This will become clear as we look at the data.

### Question 1: How common is demonstrating in European democracies?

```{r, echo = FALSE, results = 'asis'}
ess %>% 
filter(!is.na(demo)) %>% 
count(demo) %>%
  mutate(percent = round(n/sum(n)*100, digits=1)) %>%
  knitr::kable("pandoc", caption = "Demonstrating in European Democracies",
  col.names = c('Demonstrating', 'N', '%'), align="lcc")
```

Let's break this down by the six countries for which we have data:

```{r, echo = FALSE, results = 'asis'}
ess %>% 
filter(!is.na(demo)) %>% 
group_by(country) %>% 
count(demo) %>%
  mutate(percent = round(n/sum(n)*100, digits=1)) %>%
  knitr::kable(caption = "Demonstrating in European Democracies",
  col.names = c('Country', 'Demonstrating', 'N', '%'), align="lccc")
```
We can graph this to make it clear

```{r, echo = FALSE}
ess %>% 
group_by(country) %>% 
filter(!is.na(demo)) %>%
  count(demo) %>% 
mutate(prop=prop.table(n*100)) %>%
  filter(!demo=="not demonstrated") %>%
  ggplot(aes(x=reorder(country, -prop), y=prop)) +
  geom_col()+
  labs(x="", 
       y="", 
       title="Figure 1: Demonstrating by Country", 
       caption="Source: ESS 2002-2016")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()
```  

What is the age pattern of the demonstrators?

```{r, echo = FALSE}
ess %>% 
group_by(agecat) %>% 
filter(!is.na(demo), !is.na(agecat)) %>%
  count(demo) %>% 
mutate(prop=prop.table(n*100)) %>%
  filter(!demo=="not demonstrated") %>%
  ggplot(aes(x=agecat, y=prop)) +
  geom_col()+
  labs(x="", 
       y="", 
       title="Figure 2: Demonstrating by Age Category", 
       caption="Source: ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()
```  

What is the education profile of demonstrators?
```{r, echo = FALSE}
ess %>% 
group_by(educat) %>% 
filter(!is.na(demo), !is.na(educat)) %>%
  count(demo) %>% 
mutate(prop=prop.table(n*100)) %>%
  filter(!demo=="not demonstrated") %>%
  ggplot(aes(x=educat, y=prop)) +
  geom_col()+
  labs(x="", 
       y="", 
       title="Figure 3: Demonstrating by Education", 
       caption="ess 2016")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()
```  

Males only?
```{r, echo = FALSE}
ess %>% 
group_by(gender) %>% 
filter(!is.na(demo), !is.na(gender)) %>%
  count(demo) %>% 
mutate(prop=prop.table(n*100)) %>%
  filter(!demo=="not demonstrated") %>%
  ggplot(aes(x=gender, y=prop)) +
  geom_col()+
  labs(x="", 
       y="", 
       title="Figure 4: Demonstrating by Gender", 
       caption="ess 2016")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()
```  

```{r, echo = FALSE}
ess %>% 
group_by(unemp) %>% 
filter(!is.na(demo), !is.na(unemp)) %>%
  count(demo) %>% 
mutate(prop=prop.table(n*100)) %>%
  filter(!demo=="not demonstrated") %>%
  ggplot(aes(x=unemp, y=prop)) +
  geom_col()+
  labs(x="", 
       y="", 
       title="Figure 5: Demonstrating by Unemployed", 
       caption="Source: ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()
```

What does all of the above tell us about demonstrators?

Question 2: Do demonstrators participate?

```{r, echo = FALSE, results = 'asis'}
ess %>% 
filter(!is.na(demo), !is.na(vote)) %>%
  group_by(demo) %>% 
count(vote) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!vote == "did not vote") %>%
  knitr::kable(caption = "Demonstrating and Voting",
  col.names = c('Demonstrating', 'Voting', 'N', 'Percentage'),
  digits=2, align="cccc") 
```

```{r, echo = FALSE}
ess %>% 
filter(!is.na(demo), !is.na(vote)) %>%
  group_by(demo, country) %>% 
count(vote) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!vote == "did not vote") %>%
  ggplot(aes(x=demo, y=prop)) +
  geom_col()+
  labs(x="", 
       y="%", 
       title="Figure 6: Demonstrating and Voting by Country", 
       caption="Source: ESS 2002-2018")+
  theme_bw()+
  facet_wrap(~country, nrow = 3)
```

Petitioning 

```{r, echo = FALSE, results = 'asis'}
ess %>% 
filter(!is.na(demo), !is.na(petit)) %>%
  group_by(demo) %>% 
count(petit) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!petit == "not signed") %>%
  knitr::kable(caption = "Demonstrating and Petition",
  col.names = c('Demonstrating', 'Petition', 'N', 'Percentage'),
  digits=2, align="cccc")
```

Petitioning across countries by demonstrating

```{r, echo = FALSE}
ess %>% 
filter(!is.na(demo), !is.na(petit)) %>%
  group_by(demo, country) %>% 
count(petit) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!petit == "not signed") %>%
  ggplot(aes(x=demo, y=prop)) +
  geom_col()+
  labs(x="", 
       y="%", 
       title="Figure 7: Petitioning by Socialising and Country", 
       caption="ess 2016")+
  theme_bw()+
  facet_wrap(~country, nrow = 3)
```
Contacting by Demonstrating
```{r, echo = FALSE, results = 'asis'}
ess %>% 
filter(!is.na(demo), !is.na(contact)) %>%
  group_by(demo) %>% 
count(contact) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!contact == "have not contacted") %>%
  knitr::kable(caption = "Contacting and Petition",
  col.names = c('Demonstrating', 'Contacting', 'N', 'Percentage'),
  digits=2, align="cccc")
```

Contacting across countries by demonstrating
```{r, echo = FALSE, results = 'asis'}
ess %>% 
filter(!is.na(demo), !is.na(contact)) %>%
  group_by(demo, country) %>% 
count(contact) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!contact == "not contacted") %>%
  ggplot(aes(x=demo, y=prop)) +
  geom_col()+
  labs(x="", 
       y="%", 
       title="Figure 9: Contacting by Socialising and Country", 
       caption="Source: ESS 2002-2016")+
  theme_bw()+
  facet_wrap(~country, nrow = 3)
```

So what can we conclude overall about demonstrators? Does it look like it is a small group of people who demonstrate, or a more homogenous section of the population? 