---
title: "Lab 3: Political Parties"
author: "Ross Campbell"
date: '2025-09-01'
output: word_document
highlight: NULL
---


```{r setup, include=FALSE}
ess <- foreign::read.dta("https://raw.github.com/drcampbell1/Participation/master/data/ess.dta", convert.factors = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(foreign)
library(tidyverse)
library(tidytext)
library(kableExtra)
```

In the lecture, we found a particularly challenging picture of political parties. We found:

1. weakening attachments;

2. falling (and concentrating) membership; and

3. Negative public images.

In this lab, we're going to look at who works for political parties, examining their socio-economic background to see if there is inequality.

Let's begin by looking at the general levels of working for parties:

```{r, echo = FALSE, results='asis'}

ess %>% filter(!is.na(party)) %>% count(party) %>%
  mutate("%" = round(n/sum(n) * 100, digits=1)) %>%
  knitr::kable("pandoc", caption = "Work for Party in European Democracies",
  col.names = c('Worked', 'N', '%'), align="lccc")
```

How does this vary by country?
```{r, echo = FALSE, results='asis'}
ess %>% 
  group_by(country) %>% 
  filter(!is.na(party)) %>%
  count(party) %>% 
  mutate("%" = round(n/sum(n) * 100, digits=1)) %>%
  knitr::kable("pandoc", caption = "Comparing Work for Party in European Democracies",
  col.names = c('Country', 'Worked', 'N', '%'), align="cccc")
```

Let's graph this to make the findings stand out a bit better
```{r, echo = FALSE, fig.height=25, fig.width=35}

ess %>% 
  group_by(country) %>% 
  filter(!is.na(party)) %>%
  count(party) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!party=="not worked") %>%
  ggplot(aes(x=reorder(country, -prop), y=prop)) +
  geom_col()+
  labs(x="", y="", title="Figure 1: Worked for Party by Country", caption="ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 30, 
                                   hjust =1, 
                                   size = 46),
              axis.text.y = element_text(size = 46),
        plot.title = element_text(size = 46),
        plot.subtitle = element_text(size=46),
        plot.caption = element_text(size = 40))
``` 

How does the age of the population differ from that of workers of political parties?


```{r, include = FALSE, results = 'asis'}
a <- ess %>% 
  group_by(country) %>% 
  summarise(mean= mean(age, na.rm=T))
b <- ess %>% 
  group_by(country, party) %>% 
  summarise(avg_age_pp= mean(age, na.rm=T)) %>% 
  filter(!is.na(party), !party=="not worked")
left_join(a, b, by = "country") %>% select(-party) %>%
     knitr::kable("pandoc", caption = "Average Population and Workers of Parties in European Democracies",
     col.names = c('Country', 'Average Age Population', 'Average Age Worker Political Party'), digits =1, align="cccc")

rm(a,b)

```

Is there gender balance amongst workers of political parties compared with the population?
```{r, include = FALSE, results = 'asis'}
c <- ess %>% group_by(country) %>%
  filter(!is.na(gender))%>%
  count(gender) %>%
  mutate(perc_pop=n/sum(n)*100)

d <- ess %>% filter(party=="not worked") %>% 
       group_by(country, gender) %>%
       count(party) %>% 
       group_by(country) %>% 
        mutate(perc_wk_pp=n/sum(n)*100) %>%
       select(-party)

left_join(c, d, by=c("country", "gender")) %>% select(-n.x, -n.y) %>%
  knitr::kable("pandoc", caption = "Gender Balance in Population and Workers of Parties in European Democracies",
               col.names = c('Country', 'Gender', 'Population', 'Workers of Political Party'),
               digits =1, align="cccc")
rm(c,d)
```

Let's take a look at Education

```{r, echo = FALSE}

ess %>% 
  group_by(educat) %>% 
  filter(!is.na(party), !is.na(educat)) %>%
       count(party) %>% 
  mutate(prop=prop.table(n*100)) %>%
       filter(!party=="not worked") %>%
       ggplot(aes(x=educat, y=prop)) +
       geom_col()+
       labs(x="", y="", title="Figure 2: Worked for Party by Education", caption="ESS 2002-2018")+
       scale_y_continuous(labels=scales::percent)+
       theme_bw()+
  theme(axis.text.x = element_text(angle = 30, hjust =1))

```

```{r, echo = FALSE}

ess %>% 
  group_by(educat, country) %>% 
  filter(!is.na(party), !is.na(educat)) %>%
  count(party) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!party=="not worked") %>%
  ggplot(aes(x=educat, y=prop)) +
  geom_col()+
  facet_wrap(~country, nrow = 3)+
  labs(x="", y="", title="Figure 3: Worked for Party by Education and Country", caption="ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 30, hjust =1))
```

Economically Optimistic Party Workers?

```{r, echo = FALSE}
ess %>% 
  group_by(satecon) %>% 
  filter(!is.na(party), !is.na(satecon)) %>%
  count(party) %>% mutate(prop=prop.table(n*100)) %>%
  filter(!party=="not worked") %>%
  ggplot(aes(x=satecon, y=prop)) +
  geom_col()+
  labs(x="", y="", title="Figure 4: Worked for Party by Economic Optimism", caption="ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 30, hjust =1))
```  

```{r, echo = FALSE}

ess %>% 
  group_by(satecon, country) %>% 
  filter(!is.na(party), !is.na(satecon)) %>%
  count(party) %>% mutate(prop=prop.table(n*100)) %>%
  filter(!party=="not worked") %>%
  ggplot(aes(x=satecon, y=prop)) +
  geom_col()+
  facet_wrap(~country, nrow = 3)+
  labs(x="", y="", title="Figure 5: Worked for Party by Economic\nOptimism and Country", caption="ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 30, hjust =1))
```

```{r, include = FALSE}
rm(a,b,c,d)
```

What can we conclude about people who work for parties? Are they a model of equality and diversity?
