---
title: 'Lab 5: Social Capital'
author: "Ross Campbell"
date: '2025-09-01'
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(foreign)
library(tidyverse)
library(tidytext)
library("kableExtra")
ess <- foreign::read.dta("https://raw.github.com/drcampbell1/Participation/master/data/ess.dta", convert.factors = TRUE)
```

Question 1: How trusting are we?

We're going to start by looking at the variable measuring social trust. This is a question asking how trusting we are of other people and is perhaps the most widely used indicator of social capital.
```{r, echo = FALSE, results = 'asis'}
ess %>% 
  filter(!is.na(strust)) %>% 
  count(strust) %>%
  mutate(percent = round(n/sum(n)*100, digits=1)) %>%
  knitr::kable("pandoc", caption = "Social Trust in European Democracies",
  col.names = c('Social Trust', 'N', '%'), align="lccc")
```

The pattern varies, but it's possible that a lot of the variation comes from countries with particularly low levels of social trust. We need to compare the different countries within our sample. Lets take a look:

```{r, echo = FALSE, results = 'asis'}
ess %>% 
  filter(!is.na(strust)) %>% 
  group_by(country) %>%
  count(strust) %>% 
  mutate(percent = round(n/sum(n)*100, digits=1)) %>%
  knitr::kable("pandoc", 
               caption = "Social Trust in European Democracies",
              col.names = c('Country', 
                            'Socialise', 
                            'N', 
                            '%'), 
              align="lcccc")
```

And we can graph this to make it clear

```{r, echo = FALSE}
ess %>% 
  group_by(country) %>% 
  filter(!is.na(strust)) %>%
  count(strust) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!strust=="low" & !strust == "medium") %>%
  ggplot(aes(x=reorder(country, -prop), y=prop)) +
  geom_col()+
  labs(x="", 
       y="", 
       title="Figure 1: Social Trust by Country", 
       caption="Source: ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 30, hjust =1))
``` 

Has anything changed over time?

```{r, echo = FALSE}

ess %>% group_by(year, country) %>% 
  filter(!is.na(strust)) %>%
  count(strust) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!strust=="low" & !strust == "medium") %>%
  ggplot(aes(year, prop))+
  geom_line()+
  facet_wrap(~country, ncol = 4)+
  labs(x="", y="", title="Figure 2: Social Trust Over Time", caption="ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()+
  scale_x_continuous(breaks = c(2002, 2006, 2010, 2014, 2018))+
  theme(axis.text.x = element_text(angle = 30, hjust =1))
```

Do the figures differ by age category?

```{r, echo = FALSE}
ess %>% 
  group_by(agecat) %>% 
  filter(!is.na(strust), !is.na(agecat)) %>%
  count(strust) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!strust=="low" & !strust == "medium") %>%
  ggplot(aes(x=agecat, y=prop)) +
  geom_col()+
  labs(x="", 
       y="", 
       title="Figure 3: Social Trust by Age Category", caption="Source: ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()
```

Do the figures differ by Education?

```{r, echo = FALSE}

ess %>% 
  group_by(educat) %>% 
  filter(!is.na(strust), !is.na(educat)) %>%
  count(strust) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!strust=="low" & !strust == "medium") %>%
  ggplot(aes(x=educat, y=prop)) +
  geom_col()+
  labs(x="",
       y="",
       title="Figure 4: Social Trust and Education",
       caption="ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()
```  

And Gender?

```{r, echo = FALSE}
ess %>% 
  group_by(gender) %>% 
  filter(!is.na(strust), !is.na(gender)) %>%
  count(strust) %>% 
  mutate(prop=prop.table(n*100)) %>%
  filter(!strust=="low" & !strust == "medium") %>%
  ggplot(aes(x=gender, y=prop)) +
  geom_col()+
  labs(x="",
       y="",
       title="Figure 5: Social Trust and Gender",
       caption="Source: ESS 2002-2018")+
  scale_y_continuous(labels=scales::percent)+
  theme_bw()
```  

 Section 2: Does this influence participation?

Although we have established the people are relatively social, one might ask: does it matter? Does the propensity to be socially active spur participation? And does it vary across countries?

 Voting

```{r, echo = FALSE, results = 'asis'}
ess %>% 
  filter(!is.na(strust), !is.na(vote)) %>%
  group_by(strust) %>% count(vote) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!vote == "did not vote") %>%
  knitr::kable("pandoc", 
               caption = "Social Trust and Electoral Turnout",
  col.names = c('Socialising', 'Voting', 'N', 'Percentage'),
  digits=2, align="cccc")
```


```{r, echo = FALSE}

ess %>% 
  filter(!is.na(strust), !is.na(vote)) %>%
  group_by(strust, country) %>% 
  count(vote) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!vote == "did not vote") %>%
  ggplot(aes(x=strust, y=prop)) +
  geom_col()+
  labs(x="", y="%", title="Figure 6: Voting by Social Trust and Country", caption="ESS 2016")+
  theme_bw()+
  facet_wrap(~country, nrow= 3)
```

 Contacting

```{r, echo = FALSE, results = 'asis'}

ess %>% 
  filter(!is.na(strust), !is.na(contact)) %>%
  group_by(strust) %>% 
  count(contact) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!contact == "not contacted") %>%
  knitr::kable("pandoc", 
               caption = "Social Trust and Contacting Politicians",
  col.names = c('Social Trust', 'Contacting', 'N', 'Percentage'),
  digits=2, align="cccc")
```

 Contacting across countries

```{r, echo = FALSE}

ess %>% 
  filter(!is.na(strust), !is.na(contact)) %>%
  group_by(strust, country) %>% 
  count(contact) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!contact == "not contacted") %>%
  ggplot(aes(x=strust, y=prop)) +
  geom_col()+
  labs(x="", 
       y="%", 
       title="Figure 7: Contacting by Social Trust and Country", caption="Source: ESS 2002-2018")+
  theme_bw()+
  facet_wrap(~country, nrow = 3)
```

 Petition

```{r, echo = FALSE, results = 'asis'}
ess %>% 
  filter(!is.na(strust), !is.na(petit)) %>%
  group_by(strust) %>% 
  count(petit) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!petit == "not signed") %>%
  knitr::kable("pandoc", caption = "Social Trust and Petition",
  col.names = c('Social Trust', 'Petition', 'N', 'Percentage'),
  digits=2, align="cccc")
```

 Petitioning across countries by socialising

```{r, echo = FALSE}
ess %>% 
filter(!is.na(strust), !is.na(petit)) %>%
  group_by(strust, country) %>% 
  count(petit) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!petit == "not signed") %>%
  ggplot(aes(x=strust, y=prop)) +
  geom_col()+
  labs(x="", 
       y="%", 
       title="Figure 8: Petitioning by Social Trust and Country", 
       caption="Source: ESS 2002-2018")+
  theme_bw()+
  facet_wrap(~country, nrow = 3)
```

 Demonstrating

```{r, echo = FALSE, results = 'asis'}
ess %>% 
  filter(!is.na(strust), !is.na(demo)) %>%
  group_by(strust) %>% 
  count(demo) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!demo == "not demonstrated") %>%
  kbl("pandoc", caption = "Social Trust and Demonstrating",
  col.names = c('Social Trust', 'Demonstrating', 'N', 'Percentage'),
  digits=2, align="cccc")
```

Demonstrating Across Countries

```{r, echo = FALSE}
ess %>% 
  filter(!is.na(strust), !is.na(demo)) %>%
  group_by(strust, country) %>% 
  count(demo) %>%
  mutate(prop=n/sum(n)*100) %>%
  filter(!demo == "not demonstrated") %>%
  ggplot(aes(x=strust, y=prop)) +
  geom_col()+
  labs(x="", 
       y="%", 
       title="Figure 9: Demonstrating by Social Trust and Country", 
       caption="Source: ESS 2002-2018")+
  theme_bw()+
  facet_wrap(~country, nrow = 3)
```

So, time to draw some conclusions.
Does social capital matter?
